---
title: "ESM 206 Assignment 4 - Spiny Lobsters"
author: "Alex Ehrens & Courtney Krone"
date: "11/11/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
```

```{r}
library(tidyverse)
library(janitor)
library(tidyr)
library(here)
library(effsize)
library(kableExtra)
library(directlabels)
library(ggridges)
```

```{r}
lobster_abundance <- read.csv("lobster_abundance_sbc_lter.csv",
                              na = -99999) %>% 
  clean_names() %>% 
  mutate(site = as.character(site))

lobster_abundance$site[lobster_abundance$site == "IVEE"] <- "Isla Vista"
lobster_abundance$site[lobster_abundance$site == "NAPL"] <- "Naples"
lobster_abundance$site[lobster_abundance$site == "AQUE"] <- "Arroyo Quemado"
lobster_abundance$site[lobster_abundance$site == "CARP"] <- "Carpinteria"
lobster_abundance$site[lobster_abundance$site == "MOHK"] <- "Mohawk"
```

### 1. Introduction

The California spiny lobster (*Panulirus interruptus*) is an important species along the Southern California coast and plays a key role in maintaining diversity in kelp forests (National Park Service, 2016). These lobsters are a common target for fishermen, which greatly affects their size and abundance in the kelp forest ecosystems (National Park Service, 2016). This report explores the relationship between the abundance and size distribution of the California spiny lobster at 5 locations along the Santa Barbara channel. On January 1st, 2012, two of the 5 sites were designated as marine protected areas (MPAs) by the California Department of Fish and Wildlife (SBC LTER, 2019). Human activity, including fishing, is restricted in MPAs, and as a result many have shown positive trends of habitat recovery (Santa Barbara Channelkeeper). This report hopes to answer the question of whether MPAs affect the California spiny lobster size and abundance along the Santa Barbara coast. This will help assess the recovery of the kelp forest ecosystem. 

### 2. Data and Methods

The two variables being analyzed in this report are lobster size (mm) and lobster abundance at five sites along the Santa Barbara channel between years 2012 and 2018. Data on abundance and size were collected annually by divers just before the start of the fishing season in late summer (SBC LTER, 2019). Lobster counts were compared at each site over the time period of 2012 to 2018. Welch’s two sample t-test and Cohen's d effect size test were used to compare lobster sizes at MPA and non-MPA sites for 2012 and 2018. The following scenarios were compared:

- 2012 MPA vs.2012 Non-MPA sites
- 2018 MPA vs. 2018 Non-MPA sites
- 2012 MPA vs. 2018 MPA sites
- 2012 Non-MPA vs. 2018 Non-MPA sites

A 95% confidence level was used to determine significance. Rstudio Version 1.2.1335 was used to analyze data, creat graphs and tables, and run statistical tests.

### 3. Results

#### a. Results A - Lobster Abundance

```{r}
lobster_count <- lobster_abundance %>% 
  group_by(year, site) %>% 
  summarize(total_counts = sum(count)) %>% 
  mutate(Category = ifelse(site %in% c("Isla Vista", "Naples"), "MPA", "Non-MPA")) %>% 
  mutate(Site = site)
  
ggplot(lobster_count, aes(x = year, y = total_counts)) +
  geom_line(aes(color = Site, linetype = Category)) +
  theme_bw() +
  theme(legend.position = "right",
        legend.key.height = unit(6, "mm"),
        legend.key.width = unit(4, "mm"),
        plot.caption = element_text(hjust = 0),
        plot.subtitle = element_text(size = 8)) +
  scale_x_continuous(limits = c(2012, 2018),
                     expand = c(0,0)) +
  scale_y_continuous(limits = c(0, 1000),
                     expand = c(0,0)) +
  labs(x = "Year (2012-2018)",
       y = "Total lobster counts",
       title = "Spiny lobster abundance in the Santa Barbara Channel (2012-2018)",
       subtitle = "Source: SBC LTER",
       caption = "Figure 1. Counts of California spiny lobster at 5 sites in the Santa Barbara Channel between 2012 and 2018. Solid \nlines indicate the sites that are in MPAs (Isla Vista and Naples), and dashed lines are sites that are not in MPAs \n(Arroyo Quemado, Carpinteria, and Mohawk).")

```

In general, the total lobster abundance at each of the 5 survey sites have increased from 2012 to 2018. However, the only two sites to show an increase in abundance over the final year of study (2017 to 2018) are Isla Vista and Naples, both of these sites are located in Marine Protected Areas. The three sites that are not in Marine Protected Areas show a decrease in total lobster abundance over the final year of the survey. It appears from the trends in Figure 1 that designated Marine Protected Areas help improve lobster abundance in those areas.

#### b. Results B - Lobster Size Distribution

```{r}
lobster_size <- lobster_abundance %>% 
  filter(year %in% c("2012", "2018")) %>% 
  uncount(count) %>% 
  mutate(Year = as.character(year))

ggplot(lobster_size, aes(y = Year, x = size_mm, fill = Year, color = Year)) +
  geom_density_ridges(color = "gray40",
                      alpha = 0.7) +
  facet_wrap(~site) +
  theme_bw() +
  scale_y_discrete(expand = c(0,0)) +
  theme(legend.position = c(0.84, 0.25),
        plot.caption = element_text(hjust = 0),
        plot.subtitle = element_text(size = 8)) +
  labs(y = "Year",
       x = "Lobster Size (mm)",
       title = "Spiny lobster size distribution in the Santa Barbara Channel in 2012 and 2018",
       subtitle = "Source: SBC LTER",
       caption = "Figure 2. Size distributions of California spiny lobsters at 5 survey sites in the Santa Barbara Channel, collected in \n2012 and 2018. Lobster sizes (in mm) were measured at Arroyo Quemado (n = 38 in 2012, n = 54 in 2018), \nCarpinteria (n = 78 in 2012, n = 343 in 2018), Isla Vista (n = 26 in 2012, n = 946 in 2018), Mohawk (n = 83 in 2012, \nn = 164 in 2018), and Naples (n = 6 in 2012, n = 298 in 2018). Isla Vista and Naples are in MPAs, the rest are not.")
  
```

The density plots in Figure 2 show the distribution of lobster sizes (mm) at each site in 2012 and 2018. The two MPA sites, Isla Vista and Naples, show a positive shift in lobster size distribution in 2018 compared to 2012. However, the non-MPA sites, Arroyo Quemado, Mohawk and Carpinteria, show little to no shifts in lobster size distribution. Based on trends shown in Figure 2, lobster populations are able to grow larger at MPA sites which may indicate they are healthier and living longer due to reduced fishing and a flourishing habitat.

#### c. Results C - Compare Mean Lobster Sizes

```{r}
lobster_compare <- lobster_size %>% 
  mutate(site_category = if_else(site %in% c("Naples", "Isla Vista"), "MPA", "Non-MPA")) %>% 
  select(year, site, size_mm, site_category)

lobster_summary <- lobster_compare %>% 
  group_by(year, site_category) %>% 
  summarize(
    mean_size = mean(size_mm, na.rm = TRUE),
    sd_size = sd(size_mm, na.rm = TRUE),
    lobster_number = n()) 

# summary table of lobster statistics 
##### ADD CAPTION ABOVE TABLE
lobster_summary %>% 
  kable(digits = 2,
        caption = "Table 1. Comparing summary statistics (mean, standard deviation, and number of counts) of California spiny lobster size distributions at MPA vs. non-MPA sites in 2012 and 2018",
        col.names = c("Year",
                      "MPA or Non-MPA",
                      "Mean Size (mm)",
                      "Standard Deviation (mm)",
                      "Count")) %>% 
  kable_styling(bootstrap_options = "bordered",
                full_width = FALSE,
                position = "left")

#### start performing two-sample t-tests for 2012 MPA vs. non-MPA, 2018 MPA vs. non-MPA, 2012 MPA vs 2018 MPA, 2012 non-MPA vs 2018 non-MPA

# Create vectors 
lobster2012mpa <- lobster_compare %>% 
  filter(year == "2012") %>% 
  filter(site_category == "MPA") %>% 
  pull(size_mm)

lobster2012nonmpa <- lobster_compare %>% 
  filter(year == "2012") %>% 
  filter(site_category == "Non-MPA") %>% 
  pull(size_mm)

lobster2018mpa <- lobster_compare %>% 
  filter(year == "2018") %>% 
  filter(site_category == "MPA") %>% 
  pull(size_mm)

lobster2018nonmpa <- lobster_compare %>% 
  filter(year == "2018") %>% 
  filter(site_category == "Non-MPA") %>% 
  pull(size_mm)

# t-test and cohen's d comparing mpa and non-mpa sites in 2012
lobster2012_ttest <- t.test(lobster2012mpa, lobster2012nonmpa)

lobster2012_d <- cohen.d(lobster2012mpa, lobster2012nonmpa)

# t-test and cohen's d comparing mpa and non-mpa sites in 2018
lobster2018_ttest <- t.test(lobster2018mpa, lobster2018nonmpa)

lobster2018_d <- cohen.d(lobster2018mpa, lobster2018nonmpa)

# t-test comparing mpa sites in 2012 and 2018
lobster_mpa_ttest <- t.test(lobster2012mpa, lobster2018mpa)

lobster_mpa_d <- cohen.d(lobster2012mpa, lobster2018mpa)

# t-test comparing non-mpa sites in 2012 and 2018
lobster_nonmpa_ttest <- t.test(lobster2012nonmpa, lobster2018nonmpa)

lobster_nonmpa_d <- cohen.d(lobster2012nonmpa, lobster2018nonmpa)

```
##### Results C notes (DELETE LATER)
MPA vs. Non-MPA in 2012: 

- Diff. in means = (74.92 - 67.38) = 7.54
- Non-MPA > MPA
- t = -3.2516, p-value = 0.002263
- Cohen's d = -0.6098605 (medium)

MPA vs. Non-MPA in 2018:

- Diff. in means = (77.57 - 73.62) = 3.95
- MPA > Non-MPA
- t = 7.3071, p-value = 4.876e-13
- Cohen's d = 0.3514986 (small)

MPA in 2012 vs. MPA in 2018:

- Diff. in means = (77.57 - 67.38) = 10.19. 
- Increase in mean size from 2012 to 2018
- t = -4.6879, p-value = 4.76e-05
- Cohen's d = -0.8700026 (large)

Non-MPA in 2012 vs. Non-MPA in 2018:

- Diff. in means = (74.92 - 73.62) = 1.30
- Small decrease in mean size from 2012 to 2018
- t = 1.3343, p-value = 0.1831
- Cohen's d = 0.1213778 (negligible)

##### Results C write-up

MPA sites are typically chosen at spots that have the best chance of restoring the bounty of ocean life. The abundance in these areas may have attracted many fishermen to fish in that area causing lobster abundance and size to be very low in these areas.

**MPA vs. Non-MPA in 2012**

The differences between mean lobster sizes between MPA vs. Non-MPA sites in 2012 is larger with Non-MPA sites having larger mean lobster sizes and a medium effect size. A relatively large p-value of 0.002263 was also calculated. This could potentially mean that MPA sites were fished heavily compared to Non-MPA before it became a designated protected area.

**MPA vs. Non-MPA in 2018**

Null hypothesis: There is no difference in the mean lobster size for MPA vs. Non-MPA sites for 2018.
Alternative hypothesis: There is a difference in mean lobster size for MPA vs. Non-MPA sites for 2018.
The difference in means between MPA vs. Non-MPA during 2018 is 3.95 with a small effect size.  A small p-value of 4.876 x10^-13 was also calculated. This means that the probability of the null hypothesis being true is very small.

**MPA in 2012 vs. MPA in 2018**

Null hypothesis: There is no difference between the mean lobster size of MPA sites for 2012 and MPA sites in 2018.
Alternative hypothesis: There is a difference between mean lobster sizes for MPA sites for 2012 and MPA sites in 2018.
The difference in means between MPA sites in 2012 and MPA sites in 2018 is larger with a large effect size. A small p-value 4.76 x10^-5 was also calculated. This means that the probability of the null hypothesis being true is very small.

**Non-MPA in 2012 vs. Non-MPA in 2018**

Null hypothesis: There is no difference between mean lobster size of Non-MPA sites for 2012 and Non-MPA sites in 2018.
Alternative hypothesis: There is a difference between mean lobster size of Non-MPA sites for 2012 and Non-MPA sites in 2018.





### 4. References

SBC LTER (2019). https://portal.lternet.edu/nis/mapbrowse?packageid=knb-lter-sbc.77.3

National Park Service (2016). https://www.nps.gov/chis/learn/nature/california-spiny-lobster.htm

Santa Barbara Channelkeeper (no date). https://www.sbck.org/current-issues/marine-conservation/marine-protected-areas/learn-about-mpas/
